---
title: "MARSS Model to Iterpolate Missing Streamflow Observations"
author: "Thomas Buehrens"
output:
  html_document:
    code_folding: hide
    fig_caption: yes
    theme: cerulean
    toc: yes
    toc_depth: 3
    toc_float: yes
  pdf_document:
    toc: yes
    toc_depth: '3'
---

***

Last Updated `r format(Sys.time(), '%m/%d/%Y')`.

***


## Setup
All analyses require R software [**(link)**](https://cran.r-project.org/) (v3.4.3) for data retrieval, data processing, and summarizing model results. Here we configure R to perform our analysis and generate our outputs
```{r set_options, echo = TRUE, message = FALSE}
options(width = 100)
knitr::opts_chunk$set(message = FALSE)
set.seed(123)
```

We also need a couple of helper functions which we will load from the functions folder.
```{r load_funcs, message = FALSE, warning = FALSE,results = "hide"}
wd_functions<-"functions"
sapply(FUN = source, paste(here::here(),wd_functions, list.files(wd_functions), sep="/"))
```

Here we will load & install packages we need to use (needs internet connection if packages not already installed)
```{r load_packages, message = FALSE, warning = FALSE,results = "hide"}
packages_list<-c("MARSS",
                 "tidyverse",
                 "lubridate", 
                 "kableExtra",
                 "dataRetrieval"
                 ) 
install_or_load_pack(packages_list)
```

## Flow Data
Here we'll grab some streamflow data from Southwest Washington
```{r message = FALSE, warning = FALSE,results = "hide"}
naselle<-get_USGS_data(flow_site =12010000, #Naselle
                        flow_params ="discharge",
                        daily_or_instantaneous= "daily",
                        begin_date = "2002-10-01",
                        end_date = "2022-09-30"
                        )

grays<-get_DOE_data(flow_site ="25B060", #Grays
                        flow_params ="discharge",
                        daily_or_instantaneous= "daily",
                        begin_date = "2002-10-01",
                        end_date = "2022-09-30"
                        )

elochoman<-get_DOE_data(flow_site ="25C060", #Elochoman
                        flow_params ="discharge",
                        daily_or_instantaneous= "daily",
                        begin_date = "2002-10-01",
                        end_date = "2022-09-30"
                        )

abernathy<-get_DOE_data(flow_site ="25E060", #Abernathy
                        flow_params ="discharge",
                        daily_or_instantaneous= "daily",
                        begin_date = "2002-10-01",
                        end_date = "2022-09-30"
                        )

chehalis_doty<-get_USGS_data(flow_site ="12020000", #Chehalis_doty
                        flow_params ="discharge",
                        daily_or_instantaneous= "daily",
                        begin_date = "2002-10-01",
                        end_date = "2022-09-30"
                        )

grays_stg<-get_DOE_data(flow_site ="25B060", #Grays
                        flow_params ="stage_height",
                        daily_or_instantaneous= "daily",
                        begin_date = "2002-10-01",
                        end_date = "2022-09-30"
                        )


flow_dat<-naselle%>%
  dplyr::select(date,flow_cfs)%>%
  mutate(river="naselle")%>%
  bind_rows(grays%>%
              dplyr::select(date,flow_cfs)%>%
              mutate(river="grays")
              )%>%
  bind_rows(elochoman%>%
              dplyr::select(date,flow_cfs)%>%
              mutate(river="elochoman")
              )%>%
  bind_rows(abernathy%>%
              dplyr::select(date,flow_cfs)%>%
              mutate(river="abernathy")
              )%>%
  bind_rows(chehalis_doty%>%
              dplyr::select(date,flow_cfs)%>%
              mutate(river="chehalis_doty")
              )%>%
  bind_rows(grays_stg%>%
              dplyr::select(date,flow_cfs=stage_height)%>%
              mutate(river="grays_stg")
              )%>%
  group_by(river,date)%>%
  summarise(flow_cfs=mean(flow_cfs),.groups = "keep")%>%
  ungroup()
```

## Plot Flow Data
```{r message = FALSE, warning = FALSE,results = "show", fig.show=TRUE,fig.cap="Figure 1. Observed flow (cfs) by river."}
ggplot(flow_dat,aes(x=date,y = flow_cfs,color=river),alpha=0.5)+
  geom_point(size=0.3)+
  facet_wrap(~river,scales = "free_y")+
  scale_y_log10()+
  theme_bw()+
  annotation_logticks(sides = "l") 
```

## Plot Flow Anomalies
```{r message = FALSE, warning = FALSE,results = "show", fig.show=TRUE,fig.cap="Figure 2. Observed z-scored logged flow (anomalies) by river."}
flow_dat<-flow_dat%>%
  group_by(river)%>%
  mutate(l_flow_cfs=as.vector(log(flow_cfs)))%>%
  ungroup()%>%
  mutate(flow_cfs=ifelse(river=="grays" & date > ymd("2022-01-07") | river=="grays" & flow_cfs < 30 ,NA,flow_cfs),
         )%>%
  filter(!is.na(flow_cfs))%>%
  group_by(river)%>%
  mutate(l_flow_cfs=as.vector(log(flow_cfs)))
  
  # flow_dat%>%group_by(river)%>%summarize(min_zl_flow=min(zl_flow_cfs),
  #                                        min_flow=min(flow_cfs)
  #                                        )

ggplot(flow_dat,aes(x=date,y = l_flow_cfs,color=river),alpha=0.5)+
  geom_point(size=0.3)
```

## MARSS Model 
Here we use a [MARSS model](https://cran.r-project.org/web/packages/MARSS/citation.html) to interpolate missing or bad flow data.
```{r  message = FALSE, warning = FALSE,results = "show"}
marss_dat<-flow_dat%>%
  dplyr::select(river,date,l_flow_cfs)%>%
  pivot_wider(names_from = river,values_from = l_flow_cfs,id_cols=date)%>%
  arrange(date)%>%
  mutate(jdate=julian(date))%>%
  filter(year(date)>2018)

all_dates<-tibble(jdate=julian(min(marss_dat$date)):julian(max(marss_dat$date)))

marss_dat<-all_dates%>%
  left_join(marss_dat)

marss_mat<-marss_dat%>%
  dplyr::select(-c(date,jdate))%>%
  as.matrix()%>%
  t()


model=list(
 Q= "unconstrained",#"equalvarcov",
 R= diag(rep(0,nrow(marss_mat))),#"diagonal and equal","diagonal and unequal",
 U= matrix(rep(0,nrow(marss_mat)),nrow=nrow(marss_mat),1)
)
fit=MARSS(marss_mat, model=model,control=list(maxit=2000,allow.degen=T))


fitted<-t(fit$states)
colnames(fitted)<-gsub("X.","mle_",colnames(fitted))

marss_dat<-marss_dat%>%
  dplyr::select(-jdate)%>%
  bind_cols(fitted)%>%
  pivot_longer(cols=c(everything(),-date),names_to = "river")%>%
  mutate(type=ifelse(grepl("mle_",river),"mle","obs"))%>%
  mutate(river=gsub("mle_","",river))%>%
  pivot_wider(values_from =value,names_from = type)%>%
  arrange(river,date)

# scale_params<-flow_dat%>%
#   group_by(river)%>%
#   summarise(mu_l=mean(log(flow_cfs)),
#             sd_l=sd(log(flow_cfs))
#   )

marss_dat<-marss_dat%>%
  ungroup()%>%
  #left_join(scale_params)%>%
  mutate(mle_rescaled=exp(mle))%>%
  left_join(flow_dat%>%
              dplyr::select(date,river,flow_cfs)
              )

write.csv(marss_dat,"marss_dat_log.csv",row.names = F)
```

## Plot Flow Data with MARSS Predictions
```{r  message = FALSE, warning = FALSE,results = "show", fig.show=TRUE,fig.cap="Figure 3: Comparison of Observed flow vs. re-scaled MARSS fits"}
ggplot(marss_dat,aes(x=date,y=mle_rescaled,color=river))+
  geom_line()+
  geom_point(aes(y=flow_cfs),size=0.6,color="black")+
  facet_wrap(~river, scales = "free_y")+
  scale_y_log10()+
  ylab("flow (cfs)")+
  theme(axis.text.x = element_text(angle = 90))+
  theme_bw()+
  annotation_logticks(sides = "l") 
```
